---
title: "Subetting a data file"
author: "Jeff Canar, Jackie Fox, Kelsey Howard, Kritika Nayar, Maureen Satyshur"
date: "Saturday, September 26th, 2015"
output: pdf_document
classoption: landscape
---

### Objectives:

1. Learn how to create a data frame that includes a subset of variables from a larger data frame
2. Learn how to select cases based on the value of a variable (or variables) 

### Getting Started


```{r}
setwd("~/NU/RStudentLab") # set the working directory of the data file
load("gss.Rda") # load the data
```

In order to select out a subset of variables, you

1.  Specify the variables you are interested in keeping
2.  Create a vector of the variable names (or you could use column numbers)
3.  Create a new dataframe that is made using the "[ ]" notation we learned earlier

Keep in mind that when we subset a dataframe, we create a **new** dataframe (or object).  Our original data is still there.  We are simply creating a new object that is a subset of our original object.

\newpage

We can use variable names to subset a data set, keeping only those variables that we are interested in.  

```{r}
myvars <- c("marital", "health", "life") #using variable names
My.Subset <- MyData[myvars]
head(My.Subset)
```

We can also use column numbers.

```{r}
myvars2 <- c(3, 11, 12) #using column numbers
My.Subset2 <- MyData[myvars]
head(My.Subset2)
```

In order to select specific cases (rows), we use the same "[ ]" operator we learned earlier.  Note the *','* contained within the *[ ]* notation.  Placing a ',' after the selection criteria means "select the rows that meet the criteria, but take all the columns".

For example, if we are interested in taking the first 5 observations from a data table...

```{r}
# first 5 observations
My.New.Cases <- MyData[1:5,]
dim(My.New.Cases)
```

\newpage

We can also subset our data based on the value of a particular value, or values, of a particular variable.

```{r}
# based on variable values - take all persons who are divorced
Divorced.Cases<- MyData[which(MyData$marital=='Divorced'),]
dim(Divorced.Cases)
```

Finally, we can also subset our data based on the values of different variables, by stringing our command together with *and* operator, **"&"** or the *or* operator, **"|"**.

In the following examples, note the use of the "&" and "|" logical operators.

```{r}
# based on more than one variable value - take all persons who are divorced and in poor health
Divorced.Cases.Poor.Health <- MyData[which(MyData$marital == 'Divorced' & MyData$health == 'Poor'),]
dim(Divorced.Cases.Poor.Health)
```

```{r}
# based on more than one variable value - take all persons who are divorced -or- in poor health
Divorced.Cases.Or.Poor.Health <- MyData[ which(MyData$marital == 'Divorced' | MyData$health == 'Poor') , ]
dim(Divorced.Cases.Or.Poor.Health)
```

Now we can combine the subsetting variables and subsetting cases.  Lets take all the *Divorced* persons who are in *Poor* health, and look at their happiness.

```{r, fig.width=6, fig.height=4}
myvars <- c("marital", "health", "happy") #using variable names
My.Subset <- MyData[myvars]
Divorced.Cases.Poor.Health <- My.Subset[which(My.Subset$marital == 'Divorced' & My.Subset$health == 'Poor'),]
head(Divorced.Cases.Poor.Health)
Divorced.Cases.Poor.Health$happy <- droplevels(Divorced.Cases.Poor.Health$happy) # drop unused levels
plot(Divorced.Cases.Poor.Health$happy)

```

Not surprisingly, of the subset of persons who report both being "divorced" and in "Poor" health, a good majority are reporting "Not too happy".

Lets compare that to those persons who are *Married* and in *Poor* health to see who they compare.

\newpage

```{r, fig.width=6, fig.height=3}
myvars <- c("marital", "health", "happy") #using variable names
My.Subset <- MyData[myvars]
Married.Cases.Poor.Health <- My.Subset[which(My.Subset$marital == 'Married' & My.Subset$health == 'Poor'),]
head(Married.Cases.Poor.Health)
Married.Cases.Poor.Health$happy <- droplevels(Married.Cases.Poor.Health$happy) # drop unused levels
plot(Married.Cases.Poor.Health$happy)
```

So being married might mitigate some of the effects of poor health, at least in so far as the person's happiness is concerned.

&nbsp;  

### Functions used in this session: 

c()  
dim()  
which()  
head()  
droplevels()  



****  

### Practice:

1.  Create a subset of the GSS data that contains the following variables: Marital Status, Life, Poor Mental Health Days, and Depression.  Name it "My.Subset"

2.  Using the following commands, change mntlhlth from a factor level variable to a numeric variable

> **My.Subset\$mntlhlthInt <- as.numeric(levels(My.Subset$mntlhlth))[My.Subset\$mntlhlth]**


3.  Create 2 different subsets from this newly created dateframe that includes only those persons who meet the following criteria:

+ First subset (Divorced.Cases)  
> Marital = Divorced  
> Life = Routine

+ Second subset (Married.Cases)  
> Marital = Married  
> Life = Exciting

> How many cases meet the criteria for being divorced and reporting their life as "Routine" __________________?

> How many cases meet the criteria for being married and reporting their life as "Exciting" _______________?

\newpage

4.  How many persons endorsed depression from:

> the divorced group __________________________?  

> the married group ___________________________?  

5.  From each subset created above, create a barplot looking at the depression variable to compare the frequency of depression between the two groups.

6.  Use the following set of commands to look at percentages instead of raw frequency counts:

> **d <- prop.table(table(Divorced.Cases\$depress))**  
> **m <- prop.table(table(Married.Cases\$depress))**

> **barplot(d)**  
> **barplot(m)**  


> In this case, we have told R:

+ Create a **table** of frequency counts
+ Using the **prop.table** function, convert those frequency counts to proportions and store them in *d* and *m*
+ Plot those proportions