---
title: "Subetting a data file : Answers"
author: "Jeff Canar, Jackie Fox, Kelsey Howard, Maureen Satyshur"
date: "Saturday, September 26th, 2015"
output: pdf_document
classoption: landscape
---
\pagenumbering{gobble}

```{r, echo=FALSE}
setwd("~/NU/RStudentLab")
load("gss.Rda")
```

### Answers

1.  Create a subset of the GSS data that contains the following variables: Marital Status, Life, Poor Mental Health Days, and Depression.  Name it "My.Subset"

```{r}
myvars <- c("marital", "life", "mntlhlth", "depress") #using variable names
My.Subset <- MyData[myvars]
head(My.Subset)
```

2.  Using the following commands, change mntlhlth from a factor level variable to a numeric variable

> **My.Subset\$mntlhlthInt <- as.numeric(levels(My.Subset$mntlhlth))[My.Subset\$mntlhlth]**

```{r}
My.Subset$mntlhlthInt <- as.numeric(levels(My.Subset$mntlhlth))[My.Subset$mntlhlth]
```

\newpage

3.  Create 2 different subsets from this newly created dateframe that includes only those persons who meet the following criteria:

+ First subset (Divorced.Cases)  
> Marital = Divorced  
> Life = Routine

+ Second subset (Married.Cases)  
> Marital = Married  
> Life = Exciting

```{r}
Divorced.Cases <- My.Subset[which(My.Subset$marital == 'Divorced' & My.Subset$life == 'Routine'),]
dim(Divorced.Cases)
table(Divorced.Cases$life)
DivLife <- table(Divorced.Cases$life)

Married.Cases <- My.Subset[which(My.Subset$marital == 'Married' & My.Subset$life == 'Exciting'),]
dim(Married.Cases)
table(Married.Cases$life)
MarLife <- table(Married.Cases$life)
```

> How many cases meet the criteria for being divorced and reporting their life as "Routine" __________________?

> **`r DivLife[5]`** divorced persons report their life as "Routine"

> How many cases meet the criteria for being married and reporting their life as "Exciting" _______________?

> **`r MarLife[3]`** married persons report their life as "Exciting"

```{r}
table(Divorced.Cases$depress)
DivDepress <- table(Divorced.Cases$depress)

table(Married.Cases$depress)
MarDepress <- table(Married.Cases$depress)
```

4.  How many persons endorsed depression from:

> the divorced group __________________________?  

> **`r DivDepress[4]`** divorced persons reported being told they are depressed

> the married group ___________________________?  

> **`r MarDepress[4]`** married persons reported being told they are depressed

\newpage
5.  From each subset created above, create a barplot looking at the depression variable to compare the frequency of depression between the two groups.

```{r, fig.width=4, fig.height=2.5}
Divorced.Cases$depress <- droplevels(Divorced.Cases$depress) # drop unused levels
plot(Divorced.Cases$depress)

Married.Cases$depress <- droplevels(Married.Cases$depress) # drop unused levels
plot(Married.Cases$depress)
```

\newpage

6.  Use the following set of commands to look at percentages instead of raw frequency counts:

> **d <- prop.table(table(Divorced.Cases\$depress))**  
> **m <- prop.table(table(Married.Cases\$depress))**

> **barplot(d)**  
> **barplot(m)**  

```{r, fig.width=6, fig.height=3}
d <- prop.table(table(Divorced.Cases$depress)) 
m <- prop.table(table(Married.Cases$depress))

barplot(d) 
barplot(m)

```