---
title: "Session 1 - Installing, loading, understanding and using R packages : Answers"
author: "Jeff Canar, Jackie Fox, Kelsey Howard, Kritika Nayar, Maureen Satyshur"
date: "Saturday, September 26th, 2015"
output: pdf_document
classoption: landscape
---


### Session 1 : Answers

1.  Install the r package "DescTools"

> **install.packages("DescTools", dependencies=TRUE, repos="http://cran.us.r-project.org")**

2.  Load the package "DescTools" and run help to see the contents of the package

```{r}
library(DescTools)
```

3.  This package contains a data set call "d.pizza".  Load this data set and run help for the data set to see the contents.

```{r}
data(d.pizza)
```

+ How many variables are contained within the data set? ______________________________

> here are **`r ncol(d.pizza)`** variables contained within the data set.

\newpage

4.  Look at the first few cases of the data set using the "head" command

```{r}
head(d.pizza)
```

\newpage

5.  Using the "Freq" function found in "DescTools", run a frequency count for "operator"

+ How many pizza's were delivered by Maria _______________?

```{r}
Freq(d.pizza$operator)
Deliveries  <- Freq(d.pizza$operator)
```

> Maria delivered **`r Deliveries[2,2]`** pizzas.

6.  Create a basic plot for "operator"

```{r, fig.width=6, fig.height=3}
plot(d.pizza$operator)
```