---
title: "Session 2 - Essential Commands for Exploring the Structure of a datafile or R Object"
author: "Jeff Canar, Jackie Fox, Kelsey Howard, Maureen Satyshur"
date: "Saturday, September 26th, 2015"
output: pdf_document
classoption: landscape
---
\pagenumbering{gobble}

### Session 2 : Answers

```{r}
setwd("~/NU/RStudentLab")
load("gss.Rda")
MyData$weightInt <- as.numeric(levels(MyData$weight))[MyData$weight]
MyData$heightInt <- as.numeric(levels(MyData$height))[MyData$height]
```

1.  What class of object is "hivkiss" _________________________?

> The variable *hivkiss* has the class **`r class(MyData$hivkiss)`**

2.  What class of object is "id_" _____________________________?

> The variable *id_* has the class **`r class(MyData$id_)`**

\newpage

3.  Look at the summary data for weight.

```{r}
summary(MyData$weightInt)
WeightDesc <- summary(MyData$weightInt)
```

> What is the Median weight _____________________?

> The Median weight is **`r WeightDesc[3]`**

> What is the Average weight ____________________?

> The Average weight is **`r WeightDesc[4]`**

4.  Look at the summary data for Happy Marriage.

```{r}
summary(MyData$hapmar)
HapMarTable <- summary(MyData$hapmar)
```

> How many people are reporting "Not too happy" _________________________?

> **`r HapMarTable[4]`** respondents are reporting "Not too happy"

\newpage

5.  Create a simple plot for the Happy Marriage variable.

```{r, fig.width=6, fig.height=3}
MyData$hapmar <- droplevels(MyData$hapmar)
plot(MyData$hapmar)
```