---
title: "Session 5 - Data Management and Basic Plotting : Answers"
author: "Jeff Canar, Jackie Fox, Kelsey Howard, Maureen Satyshur"
date: "Saturday, September 26th, 2015"
classoption: landscape
output: pdf_document
---
\pagenumbering{gobble}

```{r, echo=FALSE}
setwd("~/NU/RStudentLab")
load("gss.Rda")
```

### Answers

1.  Read in the tab delimited data file entitled "ExamAnxiety.dat" and store the data in an object called "MyData".

```{r}
MyData <- read.table("ExamAnxiety.dat", sep="\t", header=TRUE)
```

2,  Reorder the factor level variable "GENDER" so that "Male" is the reference category.

```{r}
str(MyData$Gender)
MyData$Gender <- factor(MyData$Gender, levels = c("Male", "Female"))
str(MyData$Gender)
```

\newpage

3.  The variable "Revise" measures how many minutes a student reported revising their exam prior to turning it on.  The variable "Anxiety" is a measure of how anxious the student was about their performance on the exam.  Create a basic scatter plot, with a title, x-axis and y-axis labels looking at the association between Revision Time and Anxiety.

```{r}
plot(MyData$Anxiety, MyData$Revise,
     xlab = "Anxiety", 
     ylab = "Revision Time",
     main = "Association of Anxiety and Revision Time")
```

\newpage

4.  Create a subset of the date, plotting the association between Anxiety and Revision Time for women.  Check to verify that you have correctly subsetted the data.

```{r}
MyDataWomen <- MyData[which(MyData$Gender == 'Female'),]
head(MyDataWomen)
tail(MyDataWomen)

plot(MyDataWomen$Anxiety, MyDataWomen$Revise,
     xlab = "Anxiety", 
     ylab = "Revision Time",
     main = "Association of Anxiety and Revision Time - Women")
```

\newpage

5.  Create a subset of the date, plotting the association between Anxiety and Revision Time for men.

```{r}
MyDataMen <- MyData[which(MyData$Gender == 'Male'),]
head(MyDataMen)
tail(MyDataMen)

plot(MyDataMen$Anxiety, MyDataMen$Revise,
     xlab = "Anxiety", 
     ylab = "Revision Time",
     main = "Association of Anxiety and Revision Time - Men")
```

\newpage

6.  Put both plots together, in a single panel so that they can be viewed side-by-side.  However, to ensure both plots are more easily compared to one another, ensure that both plots have the same limits set on the x and y axis.

```{r, fig.height=3.5, fig.width=8, echo=FALSE, warning=FALSE, results="hide"}
par(mfrow=c(1,2))


plot(MyDataWomen$Anxiety, MyDataWomen$Revise,
     xlab = "Anxiety", 
     ylab = "Revision Time",
     title(main = "Association of Anxiety and Revision Time - Women", cex.main=0.75),
     xlim = c(0,100),
     ylim = c(0,100))

plot(MyDataMen$Anxiety, MyDataMen$Revise,
     xlab = "Anxiety", 
     ylab = "Revision Time",
     title(main = "Association of Anxiety and Revision Time - Men", cex.main=0.75),
     xlim = c(0,100),
     ylim = c(0,100))

```

\newpage

7.  Now add a fit line to both graphs, plot them just like you did above in question "6", and describe any differences you see.

```{r, fig.height=3.5, fig.width=8, echo=FALSE, warning=FALSE, results="hide"}
par(mfrow=c(1,2))

WomenFit <- lm(MyDataWomen$Anxiety ~ MyDataWomen$Revise)
plot(MyDataWomen$Anxiety, MyDataWomen$Revise,
     xlab = "Anxiety", 
     ylab = "Revision Time",
     title(main = "Association of Anxiety and Revision Time - Women", cex.main=0.75),
     xlim = c(0,100),
     ylim = c(0,100))
abline(WomenFit)


MenFit <- lm(MyDataMen$Anxiety ~ MyDataMen$Revise)
plot(MyDataMen$Anxiety, MyDataMen$Revise,
     xlab = "Anxiety", 
     ylab = "Revision Time",
     title(main = "Association of Anxiety and Revision Time - Men", cex.main=0.75),
     xlim = c(0,100),
     ylim = c(0,100))

abline(MenFit)
```

\newpage

8.  Which group, Men or Women, has the higher Rsq value?

```{r}
summary(WomenFit)
summary(MenFit)
```




